### 2015-05-22 doc/thesis/typedlua.tex
Improving the text and examples about table types

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.
Later in this chapter we will show that we use the syntax of records
to type modules and objects.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.

The table type \texttt{\{k:t\}} represents the type of a map
from values of type \texttt{k} to values of type \texttt{t|nil}.
Table types that represent maps always include the type \texttt{nil},
because Lua returns \texttt{nil} when we use a non-existing key to
index a table. 
The types of the keys are restricted to either base types or the
type \texttt{value} due to the statistics of the table constructor,
which we discussed in Section \ref{sec:statistics}.
Next, we show one example of table type to type a map from strings to integers:
\begin{verbatim}
    local t:{string:integer} = { foo = 1 } 
    local x:integer = t.foo           --> compile-time warning
    local y:integer? = t.bar          --> y gets nil
    local z:integer = t["bar"] or 0   --> z gets 0 
\end{verbatim}

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
\emph{unique}, otherwise it is \emph{closed}.
Any alias to \emph{unique} variables makes them \emph{open}, as we use the
tag \emph{open} to keep track of \emph{unique} variables that are aliased.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
\emph{fixed} to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags.
We explain these tags and their subtyping rules in more detail
in Chapter \ref{chap:system}.

@@

The following example shows that it is not sound to allow the
type of the fields to change:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942           --> compile-time warning
    person.lastname = 2013            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = {"firstname":string, "lastname":string}
\end{verbatim}

@@

Typed Lua always fixes \emph{unique} and \emph{open} table types
that appear in a top-level return statement.
This means that modules have \emph{fixed} table types in Typed Lua.
Typed Lua has this behavior to ensure that classes always implement
\emph{fixed} table types.
In the next section we will show that we build classes in a
similar way that we use to build modules, and we will also show
that classes should have \emph{fixed} table types to allow safe
single inheritance.

@@

After we define the module \texttt{mymath}, regardless of the adopted style,
users can use it in the standard way:
\begin{verbatim}
    local mymath = require "mymath"
    print(mymath.pow(2, 3))           --> 8
    print(mymath.pow(2, "foo"))       --> compile-time warning
\end{verbatim}

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer the type of this module.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements the given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

Lua has a mechanism for self-like (or JavaScript-like) delegation of
missing fields in a table.
After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
As we mentioned in Section \ref{sec:statistics},
Lua programmers often use this mechanism to simulate classes.
We will use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:new (x:number, y:number):self
      local s:self = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number):()
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}

>>>>

Figure \ref{fig:tables} shows that the concrete syntax of Typed Lua
table types is restricted to either the type of the empty table,
maps, arrays, or records with an optional array part.
We made this design choice due to the results that we obtained
about the usage of the table constructor, which we discussed
in Section \ref{sec:statistics}.
More precisely, those results indicated that programmers seldom
define a table constructor that is neither an empty table nor
a table that contains only literal keys.
This means that our design can type check most of the usages of
the table constructor.
Later in this chapter we will show that we use the syntax of records
to type modules and objects.
We will also show that we need the \texttt{const} modifier while
typing object-oriented code.

The table type \texttt{\{k:t\}} represents the type of a map
from values of type \texttt{k} to values of type \texttt{t|nil}.
Table types that represent maps always include the type \texttt{nil},
because Lua returns \texttt{nil} when we use a non-existing key to
index a table. 
The types of the keys are restricted to either base types or the
type \texttt{value} due to the statistics of the table constructor,
which we discussed in Section \ref{sec:statistics}.
More precisely, those results indicated that programmers seldom
use non-literal keys in the definition of a table constructor.
For typing maps, base types can type check most of the cases where
programmers use a table constructor to initialize a map,
while the type \texttt{value} is still an option to allow programmers
to type check uncommon table constructors.
This means that this restriction to key types in maps has no
impact in the usability of Typed Lua.
Next, we show one example of table type to type a map from strings to integers:
\begin{verbatim}
    local t:{string:integer} = { foo = 1 } 
    local x:integer = t.foo           --> compile-time warning
    local y:integer? = t.bar          --> y gets nil
    local z:integer = t["bar"] or 0   --> z gets 0 
\end{verbatim}

@@

Typed Lua tags a variable that holds a table type as either
\emph{unique}, \emph{open}, \emph{fixed}, or \emph{closed}.
If a variable gets its type from a table constructor then it is
\emph{unique}, otherwise it is \emph{closed}.
Any alias to \emph{unique} variables makes them \emph{open}, as we use the
tag \emph{open} to keep track of \emph{unique} variables that are aliased.
In the previous example, \texttt{person} has an \emph{unique} type
because it has no alias.
In Section \ref{sec:oop} we will show that we use the tag
\emph{fixed} to describe the type of classes.
Typed Lua also has different subtyping rules that handle these
different tags, which we explain in Section \ref{sec:subtyping}.

@@

Before discussing each one of these rules, it is worth discussing
why Typed Lua needs \emph{unique} and \emph{open} table types.
Even though both \emph{unique} and \emph{open} table types
allow the refinement of table types, we need both tags to
increase usability and to avoid some unsafe behaviors.

If we did not have \emph{unique} table types and \emph{open}
table types had the same behavior that they have now,
we would not be able to type check some safe constructions,
like in the following example:
\begin{verbatim}
    local t:{"x":integer, "y":integer?} = { x = 1, y = 2 }
\end{verbatim}

Typed Lua type checks this example because the table constructor
has an \emph{unique} table type \texttt{\{"x":1, "y":2\}}
that is consistent with the type in the annotation.
However, if the table constructor had an \emph{open} table type,
it would not type check because it does not allow us to convert
a table field from \texttt{integer} to \texttt{integer?}.
We need this kind of behavior to avoid unsafe constructions
while aliasing \emph{unique} table types, like in the follwoing
example:
\begin{verbatim}
    local t1 = { x = 1 }
    local t2:{"x":integer} = t1
    local t3:{"x":integer?} = t1      --> compile-time warning
    t3.x = nil
\end{verbatim}

Typed Lua generates a compile-time warning in this example
because aliasing \texttt{t1} changes its type from \emph{unique} to \emph{open}.
First, Typed Lua assigns an \emph{unique} table type to \texttt{t1},
but then it changes its type to \emph{open} before aliasing
\texttt{t1} to \texttt{t2}.
Even though it is safe aliasing \texttt{t1} to \texttt{t2},
it is not safe aliasing \texttt{t1} to \texttt{t3}, as it allows
removing the value that is stored in the field \texttt{x} from
\texttt{t1} through an assignment to field \texttt{x} from \texttt{t3}.
Typed Lua uses \emph{open} table types to prevent this kind of
unsafe behavior, as it would be allowed by \emph{unique} table
types due to their loose subtyping rules.

Back to the rules that Typed Lua uses to allow the refinement
of table types, the following example shows that it is not sound
to allow the type of the fields to change:
\begin{verbatim}
    local person = { firstname = "Lewis", lastname = "Reed" }
    person.middlename = "Allan"
    person.firstname = 1942           --> compile-time warning
    person.lastname = 2013            --> compile-time warning
    print("bye bye " .. person.firstname)
\end{verbatim}

@@

As an example, the following declaration defines the type \texttt{Person}
as an alias to the table type \texttt{\{"firstname":string, "lastname":string\}}
in the remainder of the current scope:
\begin{verbatim}
    local typealias Person = { "firstname":string,
                               "lastname":string }
\end{verbatim}

@@

Typed Lua always fixes \emph{unique} and \emph{open} table types
that appear in a top-level return statement.
This means that modules have \emph{fixed} table types in Typed Lua.
This behavior ensures that classes always have \emph{fixed} table types.
In the next section we will show that we build classes in a
similar way that we use to build modules, and we will also show
that classes should have \emph{fixed} table types to allow safe
single inheritance.

@@

After we define the module \texttt{mymath}, regardless of the adopted style,
users can use it in the standard way, but with the static type checking
provided by Typed Lua.
Next we show an example:
\begin{verbatim}
    local mymath = require "mymath"
    print(mymath.pow(2, 3))           --> 8
    print(mymath.pow(2, "foo"))       --> compile-time warning
\end{verbatim}

In Typed Lua, \texttt{require} is a primitive that statically type checks
a given module to infer its type.
This means that the type of its input parameter must be a literal string,
as Typed Lua uses this literal string to find the source file that
implements the given module.
To statically type check a module, Typed Lua follows the same rules
that Lua follows to load a module.
Typed Lua first checks whether the module is already statically type checked.
When the module is not yet statically type checked, Typed Lua
statically type checks its source file, and the type returned is the
type of the module.
Typed Lua raises a compile-time warning when it cannot find the source
file of a given module.

@@

Lua has a mechanism for self-like (or JavaScript-like) delegation of
missing table fields.
After the call \texttt{setmetatable(t1, \{ \string_\string_index = t2 \})},
Lua looks up in \texttt{t2} for any missing fields of \texttt{t1}.
As we mentioned in Section \ref{sec:statistics},
Lua programmers often use this mechanism to simulate classes.
We will use the following example to discuss this feature:
\begin{verbatim}
    local Shape = { x = 0.0, y = 0.0 }
    const function Shape:new (x:number, y:number):self
      local s:self = setmetatable({}, { __index = self })
      s.x = x
      s.y = y
      return s
    end
    const function Shape:move (dx:number, dy:number):()
      self.x = self.x + dx
      self.y = self.y + dy
    end
    return Shape
\end{verbatim}
